Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-03-07 young_pp_discs.tex
minor comment changes

\author[T.~A.~Douglas et al.]
{\parbox{\textwidth}{T. A. Douglas$^{1}$\thanks{E-mail: \texttt{pytd@leeds.ac.uk}},
P. Caselli$^{1}$,
J. D. Ilee$^{1}$,
A. C. Boley$^{2}$,
T. W. Hartquist$^{1}$,
R. H. Durisen$^{3}$, and
J. C. Rawlings$^{4}$\\
\vspace{0.1cm}\\
{\small{\it$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK}} \\
{\small{\it$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA}}\\
{\small{\it$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN 47405, USA}}\\
{\small{\it$^{4}$Department of Physics \& Astronomy, University College London, London WC1E 6BT, UK}}\\}}

@@

\begin{abstract}
The formation and earliest stages of protoplanetary discs are still lacking observational constraints. ALMA will soon revolutionize this field, so it is important to provide predictions and help the interpretation of future high sensitivity and high angular resolution observations. Here we present simulated ALMA observations based on radiative transfer modelling of a relatively massive (0.39 M$_{\odot}$) self-gravitating disc embedded in a 10\,M$_{\odot}$ dense core, with structure similar to the pre-stellar core L1544. We focus on simple species and conclude that  C$^{17}$O 3$\rightarrow$2, HCO$^+$ 3$\rightarrow$2, OCS 26$\rightarrow$25 and H$_2$CO 4$_{04}\rightarrow$3$_{03}$ lines together allow for measurements of the disc structure and kinematics at all scales. 
\end{abstract}

@@

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. \citealt{Krasnopolsky2011,Machida2011,Braiding2012,Joos2013}). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows. Thus, observations of the young discs, predicted to have sizes of about 100\,AU and masses as large as 10\% the original core mass (e.g. \citealt{Joos2012}), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well to filter out the extended emission tracing envelope material. \newline

After the pioneer work of, e.g., \citet{Chandler1995,Brown2000,Looney2000}, recent interferometric observations have discovered compact embedded discs in a sample of Class 0 sources \citep{Andre1999}, finding masses between 0.4 and  $>$1 M$_{\odot}$ \citep{Jorgensen2007,Jorgensen2009,Enoch2011}. A 130\,AU disc was discovered toward a Class 0 source in Perseus, using the Jansky Very Large Array (JVLA) to observe NH$_3$ \citep{Choi2007}. \citet{Pineda2012} observed methyl formate (HCOOCH$_3$) with the Atacama Large Millimetre/sub-millimetre Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disc. \citet{Persson2012} observed H$_2^{18}$O with ALMA toward the same source and found evidence of relatively large ($\ga$100\,K) excitation temperatures, as well as a HDO/H$_2$O abundance ratio close to that measured on Earth's oceans and Jupiter-family comets \citep{Hartogh2012}. \citet{Zapata2013} used ALMA to observe a disc of size $\sim$50$\,$AU in the other Class 0 source in IRAS 16293-2422 and detected infall towards this disc as inverse P-Cygni profiles in HCN and CH$_3$OH. When ALMA is completed, it will be finally possible to spatially resolve these young discs and, for the first time, put stringent constraints on the theoretical models mentioned previously. As full-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary discs. \newline

@@

To appropriately describe the environment within which a young protoplanetary disc is embedded, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (\citealt{Keto2010}, hereafter KC2010). The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere \citep{Bonnor1956,Ebert1957} with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated \citep{Caselli2012}. Figure \ref{fig:l1544_model} shows the physical parameters of the core model. \newline

@@

The pre-stellar core structure is maintained down to a radius of 80\,AU, within which the model of the young protoplanetary disc is used. The disc structure is derived by the hydrodynamic model of \citet{Boley2007} and  \citet{Boley2009}. The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating Keplerian disc (figure \ref{velocity}) featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model is sampled as a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$AU in x, y and z. The gas/dust mass ratio was assumed to be 1/100 throughout the disc and the envelope. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.\newline

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011 who followed gas-grain chemical processes during the dynamical evolution of the disc. The abundances of 125 species related by 1334 reactions were calculated through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$AU$^2$.  From I2011 we selected the four species which appear to trace different regions of the disc: the inner 20\,AU (OCS), the inner 40\,AU (H$_2$CO),  the region between $\simeq$40 and 60\,AU (HCO$^+$) and one which traces the entirety of the disc (C$^{17}$O) (see figure \ref{Chemistry}). The abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM \citep{Wilson1994}. As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also \cite{Caselli2012}). As discussed in the result section \ref{sec:model_results}, approximations have been made based on values measured toward similar objects. The molecular data used is from the LAMDA (Leiden Atomic and Molecular DAtabase) database (Sch\"oier et al. 2005; http://home.strw.leidenuniv.nl/$\sim$moldata/).

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
%-describe the RT used (LIME) 
%-describe the RT used (LIME) 
The radiative transfer program used is LIME (LIne Modeling Engine; \cite{Brinch2010}), which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in appendix  \ref{sec:gridding}. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm \citep{Lloyd1982} in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation \footnote{In three dimensions this means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figure \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in \cite{Brinch2010} to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. In order to minimise the artifacts in the output images, resulting from the grid construction, the average of ten separate runs was taken (Figure \ref{averages}).

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/grid.eps}%Lime_grid3.eps
 \caption{A plot of the points selected by the gridding process and the paths down which photons can propagate overlaid on the density distribution (in m$^{-3}$, as used in LIME). The points are more concentrated at small radii and in the densest regions.}
 \label{grid}
\end{figure}

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. In the envelope, the HCO$^+$ abundance profile follows the H$_2$O profile of L1544 \citep{Caselli2012}, scaled so the maximum is 1$\times$10$^{-8}$ (assuming, for simplicity, that photodissociation and freeze-out affect these two species in a similar way).For H$_2$CO, a step profile was considered, with a fractional abundance of 1.5$\times$10$^{-8}$ at radii greater than 0.04$\,$pc and 1.5$\times$10$^{-9}$ inside it \citep{Young2004}. For OCS we adopted a constant abundance of 1.9$\times$10$^{-9}$ \citep{Ren2011}. The C$^{17}$O abundance profile in the envelope follows the CO profile calculated for L1544 by KC2010, scaled by the 17O/16O elemental abundance.

We focus on the frequency range available with ALMA, with particular attention to band 7, which offers the best trade off between resolution and sensitivity. Thus, we consider C$^{17}$O(3$\rightarrow$2), HCO$^+$(3$\rightarrow$2), OCS (26$\rightarrow$25) and H$_2$CO(4$_{04}\rightarrow$3$_{03}$).  The frequencies and upper level energies of these transitions are in table \ref{sigmas}. The results presented in this section are limited to those lines which show detectable emission/absorption which can be used to trace either spiral structure or rotation.\newline

%We considered C$^{17}$O 3$\rightarrow$2 at 337.1$\,$GHz with an upper energy level of $E_u$=32.3K, HCO$^+$ 3$\rightarrow$2 at 267$\,$GHz and $E_u$=25.7K, OCS 26$\rightarrow$25 at 316$\,$GHz and $E_u$=205K and H$_2$CO 4$_{04}\rightarrow$3$_{03}$ at 290$\,$GHz and $E_u$=34.9K.\newline

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc) and the disc is inclined at 30$^\circ$ to edge on. From these simulated observations, integrated intensity maps, intensity weighted velocity maps and position velocity diagrams were created. The integrated intensity and intensity weighted velocity maps (figures \ref{sim_all} and \ref{mom0_maps}) were created by integrating between -12.5 to -0.5 km$\,$s$^{-1}$ and +0.5 to +12.5 km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope. This can be seen in some PV diagrams  (figure \ref{pvs}) as the strong absorption feature at all positions around zero velocity. Moment 1 maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions}.\newline

@@

OCS 26$\rightarrow$25 traces only the innermost 20$\,$AU of the disc and can be used to examine the central regions (figure \ref{sim_all}). As OCS is not seen in outflows (e.g. \citealt{Stanke2007}, \citealt{VDTak2003}) it can be used to view rotation of the central part of the disc without risk of contamination from emission from shocked material along the outflow. The OCS line, which has the highest upper energy level among the selected transitions, traces the hottest and densest regions of the disc. As a result, the innermost spiral structure is best seen in OCS (26 - 25), as shown in figure \ref{sim_all}. The majority of the OCS line is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this is towards the centre of the disc where the hole in the hot, dense mid-plane means that there is little continuum to be absorbed. However, because of the presence of the protostar and other hot material within the central hole, not considered in our simple model, we expect absorption to also be observed toward the central region of the disc.\newline

The abundance of the H$_2$CO molecule traces the spiral structure of the inner $\sim$40$\,$AU of the disc, ehich can be seen in the integrated intensity map and its rotation is detected out to larger radii than is possible with the OCS line (figure \ref{sim_all}). As with the OCS line, the majority of the H$_2$CO 4$_{04}\rightarrow$3$_{03}$, is seen in absorption against the disc mid-plane continuum, with the central region seen in emission. Unlike the OCS however, the H$_2$CO extends out to large enough radii to be present in the voids between the outer parts spiral arms. In these regions, the H$_2$CO line can be seen in emission.\newline

The fractional abundance of C$^{17}$O is constant at 2$\times$10$^{-8}$ across the disc, so that the selected C$^{17}$O line is the most accurate in reproducing the physical structure of the disc. Like H$_2$CO, C$^{17}$O shows emission in voids between spiral arms. The C$^{17}$O line is visible from the inner edge of the disc to its outer edge. Like OCS, C$^{17}$O is not seen in outflows \citep{Yildiz2012}, so we do not expect contamination.\newline

@@

In order to make predictions on the observability of the synthetic brightness maps, the Common Astronomy Software Applications package (CASA) was used to simulate their observation with the completed ALMA in the 26th most extended configuration (out of a total of 28), with the longest baseline of 14.4km and giving beam sizes between 0.02 and 0.03 arcseconds at the selected frequencies, giving a beam size close to the size of the spiral structure in the disc. As the weakest lines studied in the model are of the order of 0.1-0.2$\,$mJy$\,$beam$^{-1}$ at this resolution, sensitivities of the order of 0.02$\,$mJy$\,$beam$^{-1}$ are required. This implies around 6 hours of integration time. The exact sensitivities are given in Table \ref{sigmas}.
\begin{table*}
 \centering
 \begin{minipage}{80mm}
   \caption{Line sensitivities obtained with ALMA after 6 hours integration time with velocity resolution of $\sim$400$\,$ms$^{-1}$. As calculated by the ALMA on-line sensitivity calculator.}
   \label{sigmas}
   \begin{tabular}{c||c|c|c|c}
     \hline
     Species & Transition & Frequency & Upper energy level & Sensitivity\\
     \hline
     C$^{17}$O & 3$\rightarrow$2 & 337.06GHz & 32.3$\,$K & 0.0270$\,$mJy$\,$beam$^{-1}$ \\
     OCS & 26$\rightarrow$25 & 316.15GHz & 25.7$\,$K & 0.0275$\,$mJy$\,$beam$^{-1}$ \\
     HCO$^+$ & 3$\rightarrow$2 & 267.56GHz & 205$\,$K & 0.0176$\,$mJy$\,$beam$^{-1}$ \\
     H$_2$CO & 4$_{04}\rightarrow$3$_{03}$ &  290.62GHz & 34.9$\,$K & 0.0228$\,$mJy$\,$beam$^{-1}$ \\
     \hline
   \end{tabular}
 \end{minipage}
\end{table*}
In order to clean the images, which featured mainly absorption, the sky model used was the output of the LIME simulations with the continuum subtracted and then the spectrum inverted. The simulated images were cleaned down to a 3$\sigma$ level with $\sigma$ (the sensitivity) given in table \ref{sigmas}.\newline

@@

Figure \ref{h2co_chanmap} shows a channel map of the H$_2$CO 4$_{04}\rightarrow$3$_{03}$ line overlaid on the 1$\,$mm continuum emission. From this it can be seen that the absorption features trace the disc spiral structure whilst the line emission emanates from gaps in the outer disc, where there is less continuum to be absorbed.\newline

@@

Figure \ref{mom0_maps} shows the simulated integrated intensity maps, which demonstrates how lines from different species allow observations to trace different regions of the disc, from the OCS line tracing the innermost regions, to the HCO$^+$ line showing the outer regions which are less bright in the continuum. ALMA has enough resolution and sensitivity to see spiral structure clearly in continuum emission but also in CO isotopologues and marginally in H$_2$CO.\newline

\begin{figure*}
 \includegraphics[width=198mm]{Figures/sim/casa_all_30deg_PV_rotCurve_small.eps}
 \caption{Position velocity diagrams for each of the lines simulated with CASA. The green curve on each diagram is the best fit curve to the azimuthally averaged rotation profile from figure 4.}
 \label{pvs}
\end{figure*}

Figure \ref{pvs} shows position-velocity diagrams of the lines simulated, allowing us to reconstruct the rotation of the disc. The green curves in each panel of figure \ref{pvs} display the best fit rotation curve of the physical model (see figure \ref{velocity}), showing that the rotation profile of the disc can be well constrained using kinematic information gathered from species tracing different regions of the disc.\newline

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/casa_all_spectra2.eps}
 \caption{Simulated spectra for each of the selected transitions at the positions indicated by the coloured dots in figure \ref{mom0_maps}, the pink, green and cyan curves are taken from the locations marked the dot of the same colour respectively. In each diagram, the three spectra are taken from positions on the Y=0 axis, one is at X=0 (green) and the other 2 are from x=$\pm$ N (plus blue, minus red) where N is the distance from the centre: 10 AU for OCS, 20 AU for H$_2$CO, 28 AU for C$^{17}$O, 38 AU for HCO$^+$.}
 \label{spectra}
\end{figure*}

>>>>

\author[T.~A.~Douglas et al.]
{\parbox{\textwidth}{T. A. Douglas$^{1}$\thanks{E-mail: \texttt{pytd@leeds.ac.uk}},
P. Caselli$^{1}$,
J. D. Ilee$^{1}$,
A. C. Boley$^{2}$,
T. W. Hartquist$^{1}$,
R. H. Durisen$^{3}$, and
J. M. C. Rawlings$^{4}$\\
\vspace{0.1cm}\\
{\small{\it$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK,}} \\
{\small{\it$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA,}}\\
{\small{\it$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN 47405, USA,}}\\
{\small{\it$^{4}$Department of Physics \& Astronomy, University College London, London WC1E 6BT, UK}}\\}}

@@

\begin{abstract}
The formation and earliest stages of protoplanetary discs are still lacking observational constraints. ALMA will soon revolutionize this field, therefore it is important to provide predictions and help the interpretation of future high sensitivity and high angular resolution observations. Here we present simulated ALMA observations based on radiative transfer modelling of a relatively massive (0.39 M$_{\odot}$) self-gravitating disc embedded in a 10\,M$_{\odot}$ dense core, with structure similar to the pre-stellar core L1544. We focus on simple species and conclude that  C$^{17}$O 3$\rightarrow$2, HCO$^+$ 3$\rightarrow$2, OCS 26$\rightarrow$25 and H$_2$CO 4$_{04}\rightarrow$3$_{03}$ lines together allow for measurements of the disc structure and kinematics at all scales. 
\end{abstract}

@@

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. \citealt{Krasnopolsky2011,Machida2011,Braiding2012,Joos2013}). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows. Thus, observations of the young discs, predicted to have sizes of about 100\,au and masses as large as 10\% the original core mass (e.g. \citealt{Joos2012}), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well to filter out the extended emission tracing envelope material. \smallskip

After the pioneer work of, e.g., \citet{Chandler1995,Brown2000,Looney2000}, recent interferometric observations have discovered compact embedded discs in a sample of Class 0 sources \citep{Andre1999}, finding masses between 0.4 and  $>$1 M$_{\odot}$ \citep{Jorgensen2007,Jorgensen2009,Enoch2011}. A 130\,au disc was discovered toward a Class 0 source in Perseus, using the Jansky Very Large Array (JVLA) to observe NH$_3$ \citep{Choi2007}. \citet{Pineda2012} observed methyl formate (HCOOCH$_3$) with the Atacama Large Millimetre/sub-millimetre Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disc. \citet{Persson2012} observed H$_2^{18}$O with ALMA toward the same source and found evidence of relatively large ($\ga$100\,K) excitation temperatures, as well as a HDO/H$_2$O abundance ratio close to that measured on Earth's oceans and Jupiter-family comets \citep{Hartogh2012}. \citet{Zapata2013} used ALMA to observe a disc of size $\sim$50$\,$au in the other Class 0 source in IRAS 16293-2422 and detected infall towards this disc as inverse P-Cygni profiles in HCN and CH$_3$OH. When ALMA is completed, it will be finally possible to spatially resolve these young discs and, for the first time, put stringent constraints on the theoretical models mentioned previously. As full-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary discs. \smallskip

@@

To appropriately describe the environment within which a young protoplanetary disc is embedded, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (\citealt{Keto2010}, hereafter KC2010). The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere \citep{Bonnor1956,Ebert1957} with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated \citep{Caselli2012}. Figure \ref{fig:l1544_model} shows the physical parameters of the core model. \smallskip

@@

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure is derived by the hydrodynamic model of \citet{Boley2007} and  \citet{Boley2009}. The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating Keplerian disc (figure \ref{velocity}) featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model is sampled as a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$au in $x$, $y$ and $z$. The gas/dust mass ratio was assumed to be 1/100 throughout the disc and the envelope. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history. In the dense conditions of the protoplanetary disc could be encouraging the dust grains to agglomerate and increase the size of the ice mantle through freeze out of volitiles. This could lead to changes in the dust opacity, however we have not attempted to include these effects in our model.\smallskip

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011 who followed gas-grain chemical processes during the dynamical evolution of the disc. The abundances of 125 species related by 1334 reactions were calculated through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$au$^2$.  From I2011 we selected the four species which appear to trace different regions of the disc: the inner 20\,au (OCS), the inner 40\,au (H$_2$CO),  the region between $\simeq$40 and 60\,au (HCO$^+$) and one which traces the entirety of the disc (C$^{17}$O) (see figure \ref{Chemistry}). The abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM \citep{Wilson1994}. As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also \cite{Caselli2012}). As discussed in the result section \ref{sec:model_results}, approximations have been made based on values measured toward similar objects. The molecular data used is from the LAMDA (Leiden Atomic and Molecular DAtabase) database (Sch\"oier et al. 2005; http://home.strw.leidenuniv.nl/$\sim$moldata/).

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
The radiative transfer program used is LIME (LIne Modeling Engine; \cite{Brinch2010}), which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in appendix  \ref{sec:gridding}. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm \citep{Lloyd1982} in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation\footnote{For three dimensions, Delaunay trianglulating means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figure \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in \cite{Brinch2010} to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. In order to minimise the artifacts in the output images, resulting from the grid construction, the average of ten separate runs was taken (Figure \ref{averages}).

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime2.eps}%Lime_grid3.eps
 \caption{A plot of the points selected by the gridding process and the paths down which photons can propagate colour coded by the density distribution (in m$^{-3}$, as used in LIME). The points are more concentrated at small radii and in the most dense regions.}
 \label{grid}
\end{figure}

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. In the envelope, the HCO$^+$ abundance profile follows the H$_2$O profile of L1544 \citep{Caselli2012}, scaled so the maximum is 1$\times$10$^{-8}$ (assuming, for simplicity, that photodissociation and freeze-out affect these two species in a similar way).For H$_2$CO, a step profile was considered, with a fractional abundance of 1.5$\times$10$^{-8}$ at radii greater than 0.04$\,$pc and 1.5$\times$10$^{-9}$ inside it \citep{Young2004}. For OCS we adopted a constant abundance of 1.9$\times$10$^{-9}$ \citep{Ren2011}. The C$^{17}$O abundance profile in the envelope follows the CO profile calculated for L1544 by KC2010, scaled by the 17O/16O elemental abundance. \smallskip

We focus on the frequency range available with ALMA, with particular attention to band 7, which offers the best trade off between resolution and sensitivity. Thus, we consider C$^{17}$O(3$\rightarrow$2), HCO$^+$(3$\rightarrow$2), OCS (26$\rightarrow$25) and H$_2$CO(4$_{04}\rightarrow$3$_{03}$).  The frequencies and upper level energies of these transitions are in table \ref{sigmas}. The results presented in this section are limited to those lines which show detectable emission/absorption which can be used to trace either spiral structure or rotation.\smallskip

%We considered C$^{17}$O 3$\rightarrow$2 at 337.1$\,$GHz with an upper energy level of $E_u$=32.3K, HCO$^+$ 3$\rightarrow$2 at 267$\,$GHz and $E_u$=25.7K, OCS 26$\rightarrow$25 at 316$\,$GHz and $E_u$=205K and H$_2$CO 4$_{04}\rightarrow$3$_{03}$ at 290$\,$GHz and $E_u$=34.9K.\smallskip

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc) and the disc is inclined at 30$^\circ$ to edge on. From these simulated observations, integrated intensity maps, intensity weighted velocity maps and position velocity diagrams were created. The integrated intensity and intensity weighted velocity maps (figures \ref{sim_all} and \ref{mom0_maps}) were created by integrating between -12.5 to -0.5$\,$km$\,$s$^{-1}$ and +0.5 to +12.5$\,$km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope. This can be seen in some PV diagrams  (Figure \ref{pvs}) as the strong absorption feature at all positions around zero velocity. Moment 1 maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions}.\smallskip

@@

OCS 26$\rightarrow$25 traces only the innermost 20$\,$au of the disc and can be used to examine the central regions (figure \ref{sim_all}). As OCS is not seen in outflows (e.g. \citealt{Stanke2007}, \citealt{VDTak2003}) it can be used to view rotation of the central part of the disc without risk of contamination from emission from shocked material along the outflow. The OCS line, which has the highest upper energy level among the selected transitions, traces the hottest and densest regions of the disc. As a result, the innermost spiral structure is best seen in OCS (26 - 25), as shown in figure \ref{sim_all}. The majority of the OCS line is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this is towards the centre of the disc where the hole in the hot, dense mid-plane means that there is little continuum to be absorbed. However, because of the presence of the protostar and other hot material within the central hole, not considered in our simple model, we expect absorption to also be observed toward the central region of the disc.\smallskip

The abundance of the H$_2$CO molecule traces the spiral structure of the inner $\sim$40$\,$au of the disc, which can be seen in the integrated intensity map and its rotation is detected out to larger radii than is possible with the OCS line (figure \ref{sim_all}). As with the OCS line, the majority of the H$_2$CO 4$_{04}\rightarrow$3$_{03}$, is seen in absorption against the disc mid-plane continuum, with the central region seen in emission. Unlike the OCS however, the H$_2$CO extends out to large enough radii to be present in the voids between the outer parts spiral arms. In these regions, the H$_2$CO line can be seen in emission.\smallskip

The fractional abundance of C$^{17}$O is constant at 2$\times$10$^{-8}$ across the disc, so that the selected C$^{17}$O line is the most accurate in reproducing the physical structure of the disc. Like H$_2$CO, C$^{17}$O shows emission in voids between spiral arms. The C$^{17}$O line is visible from the inner edge of the disc to its outer edge. Like OCS, C$^{17}$O is not seen in outflows \citep{Yildiz2012}, so we do not expect contamination.\smallskip

@@

In order to make predictions on the observability of the synthetic brightness maps, the Common Astronomy Software Applications package (CASA) was used to simulate their observation with the completed ALMA in the 26th most extended configuration (out of a total of 28), with the longest baseline of 14.4km and giving beam sizes between 0.02 and 0.03 arcseconds at the selected frequencies, giving a beam size close to the size of the spiral structure in the disc. As the weakest lines studied in the model are of the order of 0.1-0.2$\,$mJy$\,$beam$^{-1}$ at this resolution, sensitivities of the order of 0.02$\,$mJy$\,$beam$^{-1}$ are required. This implies around 6 hours of integration time. The exact sensitivities are given in Table \ref{sigmas}.
\begin{table*}
 \centering
 \begin{minipage}{80mm}
   \caption{Line sensitivities obtained with ALMA after 6 hours integration time with velocity resolution of $\sim$400$\,$ms$^{-1}$. As calculated by the ALMA on-line sensitivity calculator.}
   \label{sigmas}
   \begin{tabular}{c||c|c|c|c}
     \hline
     Species & Transition & Frequency & Upper energy level & Sensitivity\\
     \hline
     C$^{17}$O & 3$\rightarrow$2 & 337.06GHz & 32.3$\,$K & 0.0270$\,$mJy$\,$beam$^{-1}$ \\
     OCS & 26$\rightarrow$25 & 316.15GHz & 25.7$\,$K & 0.0275$\,$mJy$\,$beam$^{-1}$ \\
     HCO$^+$ & 3$\rightarrow$2 & 267.56GHz & 205$\,$K & 0.0176$\,$mJy$\,$beam$^{-1}$ \\
     H$_2$CO & 4$_{04}\rightarrow$3$_{03}$ &  290.62GHz & 34.9$\,$K & 0.0228$\,$mJy$\,$beam$^{-1}$ \\
     \hline
   \end{tabular}
 \end{minipage}
\end{table*}
In order to clean the images, which featured mainly absorption, the sky model used was the output of the LIME simulations with the continuum subtracted and then the spectrum inverted. The simulated images were cleaned down to a 3$\sigma$ level with $\sigma$ (the sensitivity) given in table \ref{sigmas}.\smallskip

@@

Figure \ref{h2co_chanmap} shows a channel map of the H$_2$CO 4$_{04}\rightarrow$3$_{03}$ line overlaid on the 1$\,$mm continuum emission. From this it can be seen that the absorption features trace the disc spiral structure whilst the line emission emanates from gaps in the outer disc, where there is less continuum to be absorbed.\smallskip

@@

Figure \ref{mom0_maps} shows the simulated integrated intensity maps, which demonstrates how lines from different species allow observations to trace different regions of the disc, from the OCS line tracing the innermost regions, to the HCO$^+$ line showing the outer regions which are less bright in the continuum. ALMA has enough resolution and sensitivity to see spiral structure clearly in continuum emission but also in CO isotopologues and marginally in H$_2$CO.\smallskip

\begin{figure*}
 \includegraphics[width=180mm]{Figures/sim/casa_all_30deg_PV_rotCurve_small.eps}
 \caption{Position velocity diagrams for each of the lines simulated with CASA. The green curve on each diagram is the best fit curve to the azimuthally averaged rotation profile from figure 4.}
 \label{pvs}
\end{figure*}

Figure \ref{pvs} shows position-velocity diagrams of the lines simulated, allowing us to reconstruct the rotation of the disc. The green curves in each panel of figure \ref{pvs} display the best fit rotation curve of the physical model (see figure \ref{velocity}), showing that the rotation profile of the disc can be well constrained using kinematic information gathered from species tracing different regions of the disc.\smallskip

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/casa_all_spectra2.eps}
 \caption{Simulated spectra for each of the selected transitions at the positions indicated by the coloured dots in figure \ref{mom0_maps}, the pink, green and cyan curves are taken from the locations marked by the dot of the same colour. In each diagram, the three spectra are taken from three positions on the Y=0 axis, one is at X=0 (green) and the other 2 are from x=$\pm$ N (plus blue, minus red) where N is the distance from the centre: 10 au for OCS, 20 au for H$_2$CO, 28 au for C$^{17}$O, 38 au for HCO$^+$.}
 \label{spectra}
\end{figure*}
